---
title: "Data viz request"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyr)
require(ggplot2)
require(dplyr)
library(ggthemes)
library(scales)
```

## Importing the data

I have saved the results of this procedure in an `RData` file which will be loaded with the following code (you should rename your filepath accordingly):

```{r}
#load("~/Dropbox/R working directory/raindata workspace.RData") #linux
load("~/R working directory/raindata workspace.RData") #windows
table.original <- tabla ## creates a working table saving the original one

```

## Fixing some issues and preparing data

We see that table column names are not appropriate, fixing that

```{r}
colnames(tabla) <- c("month",tabla[1,-1]) ## assignt the column names to the first row of data
tabla <- tabla[-1,] ## deletes the first row by subsetting all rows but the first one (symbolized by the minus sign)

```

Now I will use the `gather()` function in order to transform the "wide" format into the "long" format (one row per observation). 

```{r}

my_table <- tabla %>% 
  ## adding a column for the seasons
  mutate(season = c("Summer", "Summer", "Autumn", "Autumn", "Autumn", "Winter", "Winter", "Winter", "Spring", "Spring", "Spring", "Summer")) %>%
   ## sort each monthly observation in a single row
  gather("Year","Rainfall", c(-month,-season)) %>%
  subset(select = c(2,1,3,4))

head(my_table)

```

Now I will do some more tidying up in order to R understand better the variables we are using. 

```{r}
my_table$Year <- as.integer(my_table$Year)
my_table$month <- factor(my_table$month, levels = month.name)
my_table$season <-as.factor(my_table$season)
my_table$season <- factor(my_table$season,  levels =c("Summer","Autumn","Winter","Spring"))
```

Now I'll create a new categorical variable based on the intervals of the radar values. Since the values are in mm/hr, I'll multiply each interval limit by 24. 

```{r}
clipvalues <- c(0, 4.8,   12.0,   36.0,   60.0,   96.0,  144.0,  240.0,  360.0,  480.0,  840.0, 1200.0, 1920.0, 2880.0, 4800.0, 7200.0, 8640.0)
```

The `classIntervals` function from `classInt` package uses fixed breaks. 

```{r}
require("classInt")
clippedvalues <- classIntervals(my_table$Rainfall, n = 14, style = "fixed", fixedBreaks = clipvalues)

clippedvalues
```

```{r}
my_table$catRainfall <- cut(my_table$Rainfall, clipvalues)
```

## Creating relevant visualizations

The first graph is a line graph. It uses a graphing package inside R called `ggplot`, in which each layer of a graph is inserted sequentally by a series of commands.
```{r}
# Specify which variable will go in each axis. 
by.year <- ggplot(my_table, aes(x = Year, y=Rainfall))

#Having done that, we'll start adding layers to the base plot. Each layer is added with a plus sign
by.year + 
  # stating this will be a line graph
  geom_line() + 
  # facet creates a series of graphs, one for each month
  facet_wrap(~month) +
  # assign a theme for the plot
  theme_gdocs() +
  # specify the labels of each panel
  labs(title ="Average monthly rainfall for Granpa's house", 
       x = "Year", 
       y = "Mean Rainfall (mm/day)") +
  # further customizing the graph
  theme(
    panel.background = element_rect(fill = "grey88",
                                    colour = "lightblue",
                                    size = 0.5, linetype = "solid"))

```

Here is another graph, in this case a dot chart, following the same logic of the previous graph.


```{r}

by.year + 
  geom_point(aes(color = catRainfall)) + 
  facet_wrap(~month) +
  theme_gdocs() +
  labs(title ="Average monthly rainfall for Granpa's house", 
       x = "Year", 
       y = "Mean Rainfall (mm/day)") +
  scale_color_discrete(name = "Rainfall level (mm/day)") +
  theme(
  panel.background = element_rect(fill = "grey88",
                                colour = "lightblue",
                                size = 0.5, linetype = "solid"))

```

```{r}
year_rainfall_sum <- my_table %>%
  group_by(Year) %>%
  summarise(Rainfall = sum(Rainfall, na.rm = T))
year_rainfall_sum
mean(year_rainfall_sum$Rainfall)
```


```{r}

bar_year_rainfall <- ggplot(data = year_rainfall_sum,
  aes(x = Year, y = Rainfall)) +
  # stat_summary(fun.y = sum, # adds up all observations for the month
  #   geom = "bar") +# or "line"
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks=seq(1993,2018,1),
                     expand = expand_scale(mult = c(0, 0)),
                     limits = c(1992, 2019)) +
  scale_y_continuous(breaks = seq(0,4000,500),
                     expand = expand_scale(mult = c(0.009, .05))) +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.4),
  legend.position="bottom")
bar_year_rainfall

ggsave("bar_year_rainfall.png",
       plot = bar_year_rainfall,
       width = 10.5*2,
       height = 7.425*2,
       units = "cm",
       dpi=300)


```

```{r}
bar_year_rainfall_w_avg_line <- bar_year_rainfall +
  geom_hline(yintercept = mean(year_rainfall_sum$Rainfall),
                               linetype="dashed",
                               color = "gray54", 
                               size=0.5)
bar_year_rainfall_w_avg_line
ggsave("bar_year_rainfall_w_avg_line.png",
       plot = bar_year_rainfall_w_avg_line,
       width = 10.5*2,
       height = 7.425*2,
       units = "cm",
       dpi=300)
```
```{r}
bar_year_rainfall_w_avg_line_fill <- bar_year_rainfall_w_avg_line +
  aes(fill = Rainfall > mean(year_rainfall_sum$Rainfall))+
  scale_fill_brewer(palette = "Paired",  
                    name = "", 
                    labels = c("Below average", "Above average"))
  
bar_year_rainfall_w_avg_line_fill  

ggsave("bar_year_rainfall_w_avg_line_fill.png",
       plot = bar_year_rainfall_w_avg_line_fill,
       width = 10.5*2,
       height = 7.425*2,
       units = "cm",
       dpi=300)
```

```{r}
by.month <- ggplot(my_table, 
                   aes(x = month, 
                       y=Rainfall,
                       color = season))

by.month + geom_point()
```

This graph changes the cartesian graph to a polar one, via the `coord_polar()` function. 

```{r}
## Average rainfall for each month, the whole period, colored by season
by.month + 
  coord_polar() + 
  stat_summary_bin(fun.y = "mean", 
                   geom = "bar", 
                   aes(fill = season)) 
```

Now the table has an extra column with a factor `catRainfall` that specifies to which binning the variable falls in. 

First, I'll create the palette that was extracted from that government website:
```{r}
my_palette <- c("#c9c9c9", "#B4B4FF", "#7878FF", "#1414FF", "#009690", "#006666", "#FFFF00", "#FFC800", "#FF9600", "#FF6400", "#FF0000", "#C80000", "#780000", "#280000")
```

And now the graph, 

```{r}
#Setting the base variables
discrete_graph <- ggplot(my_table, aes(x = month, y = Year))


discrete_graph + 
  coord_polar() + 
  
  # this sets the type of graph in which each month represents a tile for the period. 
  geom_tile(aes(fill = catRainfall )) +
  
  # this customize the values to the previously selected palette
  scale_fill_manual(values = my_palette, name = "Mean rainfall levels (mm/day)") +
  
  # customizing labels
  labs(title ="Average monthly rainfall for Granpa's house", 
       x = "Month", 
       y = "Year") +
  
  # further customizing
  theme(
  panel.background = element_rect(fill = "grey44",
                                colour = "lightblue",
                                size = 0.5, linetype = "solid"),
  panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
  axis.text.x=element_text(colour="grey96")
  )
  

  
```

The following chunk works for saving the last plotted graph. It won't work properly during the knitting process, so I change the `eval` value to false. 
```{r eval=FALSE}
heyplot <- last_plot()

ggsave("radial_heatmap_discrete.png", 
       plot = heyplot, 
       width = 42/4,
       height = 29.7/4,
       units = "cm",
       dpi="print")
```

## Latest modifications

```{r}
monthly_by_year <- by.year + 
  geom_point(aes(color = catRainfall),show.legend=FALSE) + 
  facet_wrap(~month) +
  theme_gdocs() +
  labs(title ="Rainfall Gauge Data [mm] at Earlville, Cairns, Australia", 
       x = "Year", 
       y = "Monthly Rainfall [mm]") +
  theme(
  panel.background = element_rect(fill = "grey88",
                                colour = "lightblue",
                                size = 0.5, linetype = "solid"),
  axis.text.x = element_text(angle = 90, 
                             hjust = 1)
  ) +
  scale_x_continuous(breaks=seq(1990,2020,5))


monthly_by_year

## Uncomment the following to write an image file to the disk (Ctrl + Shift + C on a selection does the trick)

ggsave("dot_monthly_by_year.png",
       plot = monthly_by_year,
       width = 10.5*2,
       height = 7.425*2,
       units = "cm",
       dpi=300)
```

```{r}
by.year <- ggplot(my_table, aes(x = Year, y=Rainfall))

dot_by_season <- by.year + 
  geom_point(aes(color = catRainfall),show.legend=FALSE) + 
  facet_wrap(~season,ncol = 4) +
  theme_gdocs() +
  labs(title ="Rainfall Gauge Data [mm] at Earlville, Cairns, Australia", 
       x = "Year", 
       y = "Monthly Rainfall [mm]")  +
  #scale_y_log10() +
  theme(
  panel.background = element_rect(fill = "grey88",
                                colour = "lightblue",
                                size = 0.5, linetype = "solid"),
  axis.text.x = element_text(angle = 90, 
                             hjust = 1)
  ) +
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  scale_y_continuous(breaks = seq(0,1500,250))

dot_by_season
dot_by_season_smooth <- dot_by_season + geom_smooth() 

ggsave("dot_monthly_by_season_with_smooth.png",
       plot = dot_by_season_smooth,
       width = 10.5*2,
       height = 7.425*2,
       units = "cm",
       dpi=300)

ggsave("dot_monthly_by_season_no_smooth.png",
       plot = dot_by_season,
       width = 10.5*2,
       height = 7.425*2,
       units = "cm",
       dpi=300)
```

```{r}

average_rainfall_radial <- ggplot(my_table, 
                                  aes(x=month, 
                                      y=Year)) + 
  coord_polar() + 
  geom_tile(aes(fill = Rainfall)) +
  labs(title ="Rainfall Gauge Data [mm] at Earlville, Cairns, Australia", 
       x = "Month", 
       y = "Year") 
  
  
heat_viridis <- average_rainfall_radial + 
  scale_fill_continuous(type = "viridis", 
                        name = "Monthly Rainfall [mm]") +
  scale_x_discrete(labels = toupper(month.abb)) +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        panel.background = element_rect(fill = "white")
        )
  

heat_viridis

## Uncomment the following to write an image file to the disk (Ctrl + Shift + C on a selection does the trick)

ggsave("radial_heatmap_discrete_viridis.png",
       plot = heat_viridis,
       width = 10.5*2,
       height = 7.425*2,
       units = "cm",
       dpi=300)


```

```{r}
by.year <- ggplot(my_table, aes(x = Year, y=Rainfall))

dot_by_season <- by.year + 
  geom_point(aes(color = catRainfall),show.legend=FALSE) + 
  facet_wrap(~season,ncol = 4) +
  theme_gdocs() +
  labs(title ="Rainfall Gauge Data [mm] at Earlville, Cairns, Australia", 
       x = "Year", 
       y = "Monthly Rainfall [mm]")  +
  #scale_y_log10() +
  theme(
  panel.background = element_rect(fill = "grey88",
                                colour = "lightblue",
                                size = 0.5, linetype = "solid"),
  axis.text.x = element_text(angle = 90, 
                             hjust = 1)
  ) +
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  scale_y_continuous(breaks = seq(0,1500,250))

dot_by_season
dot_by_season_smooth <- dot_by_season + geom_smooth() 

# ggsave("dot_monthly_by_season_with_smooth.png",
#        plot = dot_by_season_smooth,
#        width = 10.5*2,
#        height = 7.425*2,
#        units = "cm",
#        dpi=300)
# 
# ggsave("dot_monthly_by_season_no_smooth.png",
#        plot = dot_by_season,
#        width = 10.5*2,
#        height = 7.425*2,
#        units = "cm",
#        dpi=300)
```
